Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach

### 2023-01-24 analysis/paper/paper.Rmd
Updated manuscript

---
title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
institute:
  - UmU: Umeå University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" 
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
mainfont: Times New Roman
bibliography: references.bib
csl: "../templates/journal-of-archaeological-and-anthropological-science.csl"
abstract: |
  Northern Fennoscandia represents a geologically complex region with both glacial and postglacial processes having an impact on the archaeology. The widely distributed moraine makes provenience studies difficult, and there is a severe lack of methodological development in the characterization of local geological materials used for tool making. The question is how to approach provenience studies in a region where there is a lack of reference samples as well as documentation of known geological sources. Quartz and quartzite are two central material types utilized among hunter-gatherers in Northern Sweden, and thus important to the understanding of raw material procurement and material flow within the region. A dataset of 445 quartz and quartzite points were analyzed non-destructively with three separate spectroscopic instrumentations, including Near-Infrared, Raman scattering and X-Ray Fluorescence. The aim was to test whether non-destructive application of these spectroscopic techniques can distinguish between different quartz and quartzite types. Multivariate modelling of the spectral data was performed using Principal Component Analysis (PCA) in which the features observed in the spectra contribute to the clustering within the model. While the results demonstrate that there are a number of structural and compositional differences within the dataset it is difficult to identify any distinct material classes based on any singular technique. The features identified do, however, provide good grounds for survey efforts and the identification of potential geological origin of some of the material. Future efforts will endeavor to model the joint relations between the separate blocks in order to test whether the combined information can aid in discerning different material classes. 
keywords: |
  Spectroscopy; Quartzite; Characterization; Lithics
highlights: |
  These are the highlights. 
---

@@

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources. Flint and obsidian have both been central to the study of trade and exchange networks in European prehistory, due in part to the well-surveyed geological sources for these materials. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al. [-@prieto_etal19 p. 15]. In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region as they represent the most common material groups at archaeological sites. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp. 115-122; @lundberg97 pp. 161-171], in contrast, there has been little development on the topic of quartz and quartzite. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape. What interpretations of lithic material sourcing within Northern Sweden there are tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. Spectroscopic instrumentation feature fast measurements that can be collected directly on the surface of an object. With newer mobile models of these instruments readily available, this makes it an attractive option for field-based screening purposes. In the case of archaeological research it is also worth considering situations where destructive techniques, such as thin-section analysis, are not suitable or applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other complementary instrumentations that may provide further insight into the structural properties of a material. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

@@

When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24]. As the material frequently has been characterized based on megascopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

```{r sample-hue-pictures, fig.cap= "An assortment of points demonstrating the variation in color and heterogeneity of the material."}
knitr::include_graphics(here::here("analysis", "figures", "sample_hue.jpg"), error = FALSE)
```

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/]. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix. An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.

@@

### X-Ray Fluourescence
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Exposing a material to X-ray radiation causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e. *fluourescence*). As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can be tested statistically [@tykot03].

@@

As can be seen in the score plot (fig. \@ref(fig:xrf-pca):A), the three K-means clusters remain separated. As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (fig. \@ref(fig:xrf-kmeans-boxplot)). There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values. Loadings for elements that you would expect to follow changes in ~14~Si content also end up near cluster 1 (e.g. Al, Ca). Interestingly one of the "quartz" labelled samples ends up in cluster 3 rather than 1, and upon examining the sample visually it does not showcase any typical quartz characteristics.

@@

In the second PCA model (model B), with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%. Again, 3 clusters were deemed suitable for the K-means analysis. Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (fig. \@ref(fig:xrf-pca):B), than compared to the other clusters. The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples. Cluster 2, on the other hand, feature the main bulk of darker samples with higher ~22~Ti, ~26~Fe and ~40~Zr values. Cluster 1 likewise feature a large portion of the darker materials, with comparable ~22~Ti, ~26~Fe and ~40~Zr values. It also has higher values of ~12~Mg, ~13~Al, ~16~S, ~19~K, and ~20~Ca.

@@

```{r xrf-kmeans-boxplot, fig.cap= "Boxplot of XRF values (%) for model A, grouped by K-means cluster. Note that cluster 1 feature all those samples with a Si value above 90%, with cluster 2 featuring higher Fe/Ti/Zr values.", echo = FALSE}
knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.png"), error = FALSE)
```

@@

As quartzite is a metamorphic rock there are two factors to consider that will affect its chemical and structural properties; the make-up of the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former). The "graphite peaks" are a good indicator of such information, which could point towards the material forming in carbon-rich formations, thus helping in delimiting possible geological source areas. Usage of the term, "graphitic quartzite", can actually be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in modern day Sweden. Metamorphic rocks containing graphite, including schist and quartzite, are a common enough occurrence in other regions [e.g. @gaweda_cebulak99; @ukar_cloos16]. Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp. 482-ff], it would not be strange to encounter carbon structures in quartzite or quartzite-like material. The metal-OH features in the NIRS data is another likely piece of evidence which could be helpful in separating quartzite based on metamorphic grade.

@@

<!--

Although there are trends that align with the perceived color of the objects there are also frequent exceptions to this rule. It is also worth mentioning that the color classification does not take the heterogeneous nature of the material into consideration.

The heterogenous nature of the material also raises the issue of sampling. While NIRS provides a somewhat less subjective approximation of the color, the limited integration surface could result in a biased represenation of the material as a whole. This is true for all of the instrumentations used in the current study, especially for Raman and XRF where the spot size is far smaller than compared to the ASD contact probe. A significant number of points in the dataset feature veins of mineral inclusions running throughout the material, making it difficult to get a representative reading using probes. Such issues could be avoided using spectral cameras where the entire surface might be captured, but it would also defeat the purpose of using field-based methods.

NIR 
- absorbance
- small metal-OH differences
- material with translucence can still have dark impurities -> higher absorbance and placed next to black matrix
- difficulties in sampling lead to issues with normalization methods? Or just material itself?

Raman
- graphite indicating

XRF
- clay minerals (K, Ca, Al)
- Higher iron (dark samples)
- High/Low Si

Sampling 
-

 While no extreme outliers showed up in the analysis there were some instances where objects were located contrary to the observed trend. The two examples of points/preforms with a dark hue in the Raman model is a good example of this. It highlights the complexity of the material and

Taking the sampling of the XRF data into consideration, some interesting patterns did emerge from the analysis. The first being the bimodal distribution of ~14~Si

The exploratory approach presented in this paper holds potential for enabling provenance studies within geologically complex regions. Screening a large enough dataset makes it possible to look at in-group variations and isolate spectroscopic distinct fingerprints [e.g. @sciuto_etal19]. These can then be used as the basis for later classification of the material into structurally different groups. The characterization of quartz and quartzite is fundamental if questions of resource procurement, trade and exchange and human mobility are to be developed within Northern Sweden. Being able to distinguish between different types of quartz and quartzite would enable the spatial analysis and modelling of distribution patterns, regardless of whether the geological source is known or not. In the current case study modelling the results of single instrumentation may have been limited for the purpose of classification, but modelling the joint relations between the separate data blocks may provide further aid in isolating distinct types of quartz and quartzite.

-->
 \linebreak

@@

The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification. Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials. The study provides an example of a non-destructive and exploratory approach to the characterization of petrographic material where there also is an absence of reference material. Not only is there a need for methodological development in the study of quartzite in archaeology, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources. Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined for information, and ultimately used to produce predictive models which can be applied to spatiotemporal questions of raw material management.

@@

@@

```{r xrf-pca-loadings-scree, fig.cap= "PCA Loadings and eigenvalues for the XRF data. a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.", echo = FALSE}
knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.png"), error = FALSE)
```
\newpage

>>>>

---
title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
institute:
  - UmU: Umeå University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" 
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
mainfont: Times New Roman
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl"
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. This leads to a severe lack of methodological development in the characterization of these same materials in an archaeological context. Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region. 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). The data was evaluated with exploratory analysis using Principal Component Analysis. The results demonstrate that detectable differences exist within the dataset for each instrumentation. A number of features identified, such as the presence of graphite, can likely be related to diagenetic/paragenetic differences in the material. Not only should these results aid in future surveys of geological sources in the region, but they will also aid in the construction of predictive models for classifying quartzite.  
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
highlights: |
  These are the highlights. 
---

@@

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources. Flint and obsidian have both been central to the study of trade and exchange networks in European prehistory, due in part to the well-surveyed geological sources for these materials. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al. [-@prieto_etal19 p. 15]. In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region as they represent the most common material groups at archaeological sites. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp. 115-122; @lundberg97 pp. 161-171], in contrast, there has been little development on the topic of quartz and quartzite. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the current landscape. Existing theories on lithic material sourcing within the North Swedish region tend to describe people employing strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. Spectroscopic instrumentation features fast measurements that can be collected directly on the surface of an object. With newer mobile instrumentations readily available, this makes it an attractive option for field-based screening purposes. In the case of archaeological research it is also worth considering situations where destructive techniques, such as thin-section analysis, are not suitable or applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other complementary instrumentations that may provide further insight into the structural properties of a material. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

# Aims

@@

When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24]. As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

```{r sample-hue-pictures, fig.cap= "Not only is the material heterogeneous in terms of colour, but the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation."}
knitr::include_graphics(here::here("analysis", "figures", "figure_2.png"), error = FALSE)
```

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/]. The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens. Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. When sampling materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix. An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the representativeness of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.

@@

### X-Ray Fluourescence
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Fluorescence produced when a material is exposed to Moseley's law [@moseley1913] recognizes that

Exposing a material to high energy particles (e.g. X-Ray photons) will excite atoms as bound electrons are ejected upon colliding with the previous particles. In order for the atom to reach its ground state again one of the electrons in a higher energy level orbital will need to move to a lower energy orbital. The energy of the fluorescence produced due to this transition is characteristic for

causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e. *fluourescence*). As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can be tested statistically [@tykot03].

@@

As can be seen in the score plot (fig. \@ref(fig:xrf-pca):A), the three K-means clusters remain separated. As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%). There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values. Loadings for elements that you would expect to follow changes in ~14~Si content also end up near cluster 1 (e.g. Al, Ca). Interestingly one of the "quartz" labelled samples ends up in cluster 3 rather than 1, and upon examining the sample visually it does not showcase any typical quartz characteristics.

@@

In the second PCA model (model B), with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%. Again, 3 clusters were deemed suitable for the K-means analysis (fig. \@ref(fig:xrf-kmeans-boxplot)). Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (fig. \@ref(fig:xrf-pca):B), than compared to the other clusters. The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples. Cluster 2, on the other hand, feature the main bulk of darker samples with higher ~22~Ti, ~26~Fe and ~40~Zr values. Cluster 1 likewise feature a large portion of the darker materials, with comparable ~22~Ti, ~26~Fe and ~40~Zr values. It also has higher values of ~12~Mg, ~13~Al, ~16~S, ~19~K, and ~20~Ca.

@@

```{r xrf-kmeans-boxplot, fig.cap= "Boxplot of XRF values (%) for model B, grouped by K-means cluster. Cluster 1 is made up of samples with lower values of Si, but higher values for those elements that are typically associated with the same (e.g. Al, Ca). Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content.", echo = FALSE}
knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.png"), error = FALSE)
```

@@

As quartzite is a metamorphic rock there are two factors to consider that will affect its chemical and structural properties; the make-up of the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former). The "graphite peaks" are a good indicator of such information, which could point towards the material forming in carbon-rich formations, thus delimiting possible geological source areas. Usage of the term "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in modern day Sweden. Metamorphic rocks containing graphite, including schist and quartzite, are a common enough occurrence in other regions [e.g. @gaweda_cebulak99; @ukar_cloos16]. Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp. 482-ff], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.

Questions of metamorphic grade can also be raised based on the metal-OH features in the NIRS data. Should these be related to the Al content of the material and the presence of mica, the band positions may indicate differing metamorphic grades in the material (muscovite - biotite). Not only does such information provide grounds for discussions on provenance, but it also aids in the construction of a predictive model. The main issue, however, seems to be the struggle with sampling dark material using NIRS (and to an extent Raman), which also feature many of the indicators for biotite in the material. Sampling in general is a core issue, as the question becomes how representative the results really are of the material/objects in question? This does not simply extend to the measurements, however, but also to the visual characterization. The division of a heterogeneous material into larger simplified color groups might give a mistaken impression of homogeneity. Should a light grey point with streaks of black mineral running through it be classified as a light or dark quartzite? The necessity for adopting such classifications in the past is understandable, a certain level of abstraction is necessary in higher level modelling after all [@sjolander22]. Spectroscopic screening, however, does provide a more quantitative and empirical approach to such tasks [@ramacciotti_etal22].

@@

@@

@@

The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification. Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials. The study provides an example of a non-destructive and exploratory approach to the characterization of petrographic material where there also is an absence of reference material. Not only is there a need for methodological development in the study of quartzite in an archaeological context, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources. Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined for information, and ultimately used to produce predictive models which can be applied to spatiotemporal questions of raw material management.

@@

```{r xrf-pca-loadings-scree, fig.cap= "PCA Loadings and eigenvalues for the XRF data. a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.", echo = FALSE}
knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.png"), error = FALSE)
```
